A Data-Driven Framework for Neural Field Modeling

### 2011-01-03 ltx/first_order_framework/response_to_reviews.tex

    \section{Reviewer 2}

    We thank the author for their encouraging review. Each point raised is dealt with in turn.

\begin{enumerate}
    \item The handling of the (to borrow a term from numerical weather forecasting) `sub-grid scale physics' is inadequate for nervous systems (see, e.g., equation 8 [The nonlinear IntegroDifference Equation]). I do very much appreciate the spectral estimation of spatial and temporal scales. But the model errors are lumped into an iid model error term that is not realistic. For instance, the flow of potassium in the extracellular spaces has a known dynamic that will affect network excitability, generating important sub-grid dynamics that is anything but iid. See, for instance, the example in Ullah et al PLoS Comp Biol 2010 where without such dynamics, state tracking and parameter estimation for a neural system fails. A beautiful overview of sub-grid effects in data assimilation is found in Kalnay's 2003 book. I would very much value the authors comments on these issues. I guess were I to respond to this question, that I might consider extending the model error term to an autoregressive model. But I will let the authors answer this question.\\

	\emph{Indeed, there will always be a mismatch between a mathematical model and the natural phenomena it is attempting to describe. The model used to develop our estimator is modest in terms of the temporal dynamics, where the state transition is assumed to be first-order. However, the development of the framework for this modest brain model is a significant development that can be expanded to include longer range time dependencies. A major benefit in using a mean field model is the freedom to lump detailed parameters, such as specific ion concentrations, into more general parameters such as the connectivity kernel coefficients that describe the net system gain. Please note, we have assumed that factors influencing excitability are stationary over the estimation period. This is a reasonable assumption if applying the framework to an evoked-potential experiment. An extension to our framework, that is currently in development, is to allow for time-varying parameter estimates. This will allow the application of the framework to longer-term data recordings. The problem can then be posed in terms of convergence time and estimation accuracy vs. nonstationarity of the metabolic processes and other unmodeled brain dynamics.}

### 2011-01-03 ltx/first_order_framework/LaTeX_Manuscript.tex
edits to paper - modeling assumptions

	&= \mathbf{E}\left[m(\mathbf{r} + \boldsymbol{\tau}) \ast m(\mathbf{r}) \ast v^2(\mathbf{r}) \ast \left(\xi  + T_s \varsigma w(\mathbf{r})\right) \right] \\
	&= \left(R_{y_t,y_t}(\boldsymbol{\tau}) - \delta(\boldsymbol{\tau}\right)\sigma_{\varepsilon}^2) \ast \left(\xi  + T_s \varsigma w(\mathbf{r})\right) \label{eq:xcorr_with_autocorr}

@@

where $R_{y_{t},y_{t}}(\boldsymbol{\tau})$ is the auto-correlation of the observations and $\delta(\cdot)$ is the Kronecker delta function.

To show how we get this result, we define the auto-correlation of the observations as
\begin{align}
	R_{y_{t},y_{t}}(\boldsymbol{\tau}) &= \mathbf{E}\left[y_{t}\left(\mathbf{r}+\boldsymbol{\tau}\right)y_{t}\left(\mathbf{r}\right)\right]\\
	&= \mathbf{E}\left[m(\mathbf{r} + \boldsymbol{\tau}) \ast v(\mathbf{r}) \times m(\mathbf{r}) \ast v(\mathbf{r})\right] + \delta(\boldsymbol{\tau})\sigma_{\varepsilon}^2 \\
	&= \mathbf{E}\left[\mathcal{F}^{-1}\left\{ \exp(i\boldsymbol{\nu}^{\top}\boldsymbol{\tau})M(\boldsymbol{\nu})V(\boldsymbol{\nu}) \ast M(\boldsymbol{\nu})V(\boldsymbol{\nu}) \right\}\right] + \delta(\boldsymbol{\tau})\sigma_{\varepsilon}^2 \\
	&= \mathbf{E}\left[\mathcal{F}^{-1}\left\{ \exp(i\boldsymbol{\nu}^{\top}\boldsymbol{\tau})M(\boldsymbol{\nu})M(\boldsymbol{\nu})V(\boldsymbol{\nu}) \ast V(\boldsymbol{\nu}) \right\}\right] + \delta(\boldsymbol{\tau})\sigma_{\varepsilon}^2 \\
	&= \mathbf{E}\left[m(\mathbf{r}+\boldsymbol{\tau}) \ast m(\mathbf{r}) \ast v^2(\mathbf{r})\right] + \delta(\boldsymbol{\tau})\sigma_{\varepsilon}^2
\end{align}

@@

	\mathcal{F}\left\{R_{y_{t},y_{t+1}}(\boldsymbol{\tau})\right\} &\approx \mathcal{F}\left\{ \left(R_{y_t,y_t}(\boldsymbol{\tau}) - \delta\left(\boldsymbol{\tau}\right)\sigma_{\varepsilon}^2\right) \ast \left(\xi  + T_s \varsigma w(\mathbf{r})\right) \right\} \\
	&= \mathcal{F}\left\{ R_{y_t,y_t}(\boldsymbol{\tau}) - \delta\left(\boldsymbol{\tau}\right)\sigma_{\varepsilon}^2 \right\} \mathcal{F}\left\{ \xi  + T_s \varsigma w(\mathbf{r}) \right\}.

@@

Now rearranging

@@

	\frac{\mathcal{F}\left\{R_{y_{t},y_{t+1}}(\boldsymbol{\tau})\right\}}{\mathcal{F}\left\{ R_{y_t,y_t}(\boldsymbol{\tau}) - \delta\left(\boldsymbol{\tau}\right)\sigma_{\varepsilon}^2 \right\}} \approx \mathcal{F}\left\{ \xi  + T_s \varsigma w(\mathbf{r}) \right\}.

@@

	\mathcal{F}^{-1}\left\{\frac{\mathcal{F}\left\{R_{y_{t},y_{t+1}}(\boldsymbol{\tau})\right\}}{\mathcal{F}\left\{ R_{y_t,y_t}(\boldsymbol{\tau}) - \delta\left(\boldsymbol{\tau}\right)\sigma_{\varepsilon}^2 \right\}}\right\} &\approx \xi  + T_s \varsigma w(\mathbf{r}) % \\

@@

Let the spectral representation of the post-synaptic membrane voltage field at time $t$ be denoted by $V_t(\boldsymbol{\nu})$, where $\boldsymbol\nu$ is the spatial frequency. According to Shannon's sampling theorem, the field must be spatially band-limited for an accurate reconstruction using spatially discrete observations. However, an approximate reconstruction can be obtained if the field is approximately band-limited with 
\begin{equation}
	V_t(\boldsymbol{\nu}) \approx 0 ~ \ \forall \boldsymbol{\nu} > \boldsymbol{\nu}_c,
\end{equation}
where $\boldsymbol{\nu}_c$ is a cutoff frequency (typically taken as the -3~dB point). Given such a band-limited field, the distance between adjacent sensors, $\Delta_y$, must satisfy 
\begin{equation}
	\label{eq:MinimumSensorDistance} \Delta_y \leq \frac{1}{2\rho_y\boldsymbol{\nu}_{c}}, 
\end{equation}
where $\rho_y \in \mathbb{R} \ge 1$ is an oversampling parameter. \dean{(reviewer 1, additional comment 3:) A derivation of the sampling theorem is provided in \ref{App:SamplingTheorem} that is based on the work of \citet{Sanner1992} and \citet{Scerri2009}}. This condition must be satisfied to avoid spatial spectral aliasing effects when reconstructing the hidden dynamic field, $v_t(\mathbf{r})$, using the sampled observations, $\mathbf{y}_t$.

@@

\subsection{Modeling Assumptions}
\dean{(comment 1 and 9 from reviewer 2:) The implementation of the framework has been described a being appropriate for an evoked-potential experimental protocol. Over the time duration of an experimental paradigm the connectivity and synaptic kernel parameters are assumed to be stationary. However, if the method is applied to a continuous recording over a longer time duration it is unlikely that the parameters will be stationary. If fact, it is likely that changes in the parameter values would be of most interest.}

>>>>

	&= \mathbf{E}\left[m(\mathbf{r} + \boldsymbol{\tau}) \ast m(\mathbf{r}) \ast v^2(\mathbf{r}) \ast \left(\xi\delta_D(\mathbf{r})  + T_s \varsigma w(\mathbf{r})\right) \right] \\
	&= \left(R_{y_t,y_t}(\boldsymbol{\tau}) - \delta_K(\boldsymbol{\tau}\right)\sigma_{\varepsilon}^2) \ast \left(\xi\delta_D(\mathbf{r})  + T_s \varsigma w(\mathbf{r})\right) \label{eq:xcorr_with_autocorr}

@@

where $R_{y_{t},y_{t}}(\boldsymbol{\tau})$ is the auto-correlation of the observations, $\delta_K(\cdot)$ is the Kronecker delta function, and $\delta_D(\cdot)$ is the Dirac delta function.

To show how we get this result, we define the auto-correlation of the observations as
\begin{align}
	R_{y_{t},y_{t}}(\boldsymbol{\tau}) &= \mathbf{E}\left[y_{t}\left(\mathbf{r}+\boldsymbol{\tau}\right)y_{t}\left(\mathbf{r}\right)\right]\\
	&= \mathbf{E}\left[m(\mathbf{r} + \boldsymbol{\tau}) \ast v(\mathbf{r}) \times m(\mathbf{r}) \ast v(\mathbf{r})\right] + \delta_K(\boldsymbol{\tau})\sigma_{\varepsilon}^2 \\
	&= \mathbf{E}\left[\mathcal{F}^{-1}\left\{ \exp(i\boldsymbol{\nu}^{\top}\boldsymbol{\tau})M(\boldsymbol{\nu})V(\boldsymbol{\nu}) \ast M(\boldsymbol{\nu})V(\boldsymbol{\nu}) \right\}\right] + \delta_K(\boldsymbol{\tau})\sigma_{\varepsilon}^2 \\
	&= \mathbf{E}\left[\mathcal{F}^{-1}\left\{ \exp(i\boldsymbol{\nu}^{\top}\boldsymbol{\tau})M(\boldsymbol{\nu})M(\boldsymbol{\nu})V(\boldsymbol{\nu}) \ast V(\boldsymbol{\nu}) \right\}\right] + \delta_K(\boldsymbol{\tau})\sigma_{\varepsilon}^2 \\
	&= \mathbf{E}\left[m(\mathbf{r}+\boldsymbol{\tau}) \ast m(\mathbf{r}) \ast v^2(\mathbf{r})\right] + \delta_K(\boldsymbol{\tau})\sigma_{\varepsilon}^2
\end{align}

@@

	\mathcal{F}\left\{R_{y_{t},y_{t+1}}(\boldsymbol{\tau})\right\} &\approx \mathcal{F}\left\{ \left(R_{y_t,y_t}(\boldsymbol{\tau}) - \delta_K\left(\boldsymbol{\tau}\right)\sigma_{\varepsilon}^2\right) \ast \left(\xi\delta_D(\mathbf{r})  + T_s \varsigma w(\mathbf{r})\right) \right\} \\
	&= \mathcal{F}\left\{ R_{y_t,y_t}(\boldsymbol{\tau}) - \delta_K\left(\boldsymbol{\tau}\right)\sigma_{\varepsilon}^2 \right\} \mathcal{F}\left\{ \xi\delta_D(\mathbf{r})  + T_s \varsigma w(\mathbf{r}) \right\}.

@@

Now rearranging (Note: this step is wrong. The RHS should be a function of $\boldsymbol\tau$!)

@@

	\frac{\mathcal{F}\left\{R_{y_{t},y_{t+1}}(\boldsymbol{\tau})\right\}}{\mathcal{F}\left\{ R_{y_t,y_t}(\boldsymbol{\tau}) - \delta_K\left(\boldsymbol{\tau}\right)\sigma_{\varepsilon}^2 \right\}} \approx \mathcal{F}\left\{ \xi\delta_D\left(\mathbf{r}\right)  + T_s \varsigma w(\mathbf{r}) \right\}.

@@

	\mathcal{F}^{-1}\left\{\frac{\mathcal{F}\left\{R_{y_{t},y_{t+1}}(\boldsymbol{\tau})\right\}}{\mathcal{F}\left\{ R_{y_t,y_t}(\boldsymbol{\tau}) - \delta_K\left(\boldsymbol{\tau}\right)\sigma_{\varepsilon}^2 \right\}}\right\} &\approx \xi\delta_D\left(\mathbf{r}\right)  + T_s \varsigma w(\mathbf{r}) % \\

@@

Let the spectral representation of the post-synaptic membrane voltage field at time $t$ be denoted by $V_t(\boldsymbol{\nu})$, where $\boldsymbol\nu$ is the spatial frequency. According to Shannon's sampling theorem, the field must be spatially band-limited for an accurate reconstruction using spatially discrete observations. However, an approximate reconstruction can be obtained if the field is approximately band-limited with 
\begin{equation}
	V_t(\boldsymbol{\nu}) \approx 0 ~ \ \forall \boldsymbol{\nu} > \boldsymbol{\nu}_c,
\end{equation}
where $\boldsymbol{\nu}_c$ is a cutoff frequency (typically taken as the -3~dB point). Given such a band-limited field, the distance between adjacent sensors, $\Delta_y$, must satisfy 
\begin{equation}
	\label{eq:MinimumSensorDistance} \Delta_y \leq \frac{1}{2\rho_y\boldsymbol{\nu}_{c}}, 
\end{equation}
where $\rho_y \in \mathbb{R} \ge 1$ is an oversampling parameter. \dean{(reviewer 1, additional comment 3:) A derivation of the sampling theorem that is based on the work of \citet{Sanner1992} and \citet{Scerri2009} is provided in \ref{App:SamplingTheorem}}. The condition in equation~\ref{eq:MinimumSensorDistance} must be satisfied to avoid spatial spectral aliasing effects when reconstructing the hidden dynamic field, $v_t(\mathbf{r})$, using the sampled observations, $\mathbf{y}_t$.

@@

\subsection{Modeling Assumptions}
\dean{(comment 1 and 9 from reviewer 2:) It is an implicit assumption that the neural field model (in equation~\ref{DiscreteTimeModel}) provides an apt description of cortical dynamics. Although models of this form are capable of describing a variety of cortical dynamics, there is without doubt, a mismatch between the cortex and the model. Nevertheless, there is a major benefit in using a mean field model over a more detailed model is that the myriad of parameters that might influence excitability, such as specific ion concentrations, are considered to be lumped into the connectivity kernel coefficients that effectively describe the net system gain. Analytic solutions... A goal of the framework is to create a method to assimilate ...}

\dean{The framework assumes that the model parameters are stationary over the estimation period. This is a reasonable assumption if applying the framework to an evoked-potential experiment over a short time period in a controlled environment. However, it is well known that factors that influence excitability are non-stationary over longer time periods. Therefore, the estimation framework must be altered if it is to be applied to data over extended periods of time. A possible way to achieve this is to use a more sophisticated model.. Alternatively, a simpler method would be to allow for time-varying parameter estimates...  }

\dean{The implementation of the framework has been described a being appropriate for an evoked-potential experimental protocol. Over the time duration of an experimental paradigm the connectivity and synaptic kernel parameters are assumed to be stationary. However, if the method is applied to a continuous recording over a longer time duration it is unlikely that the parameters will be stationary. If fact, it is likely that changes in the parameter values would be of most interest.}
